
%TODO: add abstracts

\abstractEn{
In an era of an ever-increasing use of software technologies
for almost any everyday task, the requirement for high quality and flawless software is highly
indispensable. To that end, several tools leveraging program analysis techniques have been
introduced in the literature, mainly aiming to assist developers in performance optimization or bug finding.

Those tools may either be highly autonomous or possibly rely on external frameworks to back their reasoning.
A symbolic execution tool stands as an example that is highly dependent to an
external constraint solver, else known as theorem prover, that aids it by inducing a solution to an input
constraint formula which further directs the execution. Those tools essentially try to yield proofs, yet they
are implemented as complex software systems in a common programming language, not being easily extensible.
Declarative programming languages are increasingly becoming more preferable over the last few years by
the Computer Science community, with applications that span from program analysis to databases and distributed
systems. The main power of declarative languages is that the developer needs to focus on what, rather than
how he wants to achieve anything.

Motivated by the success of theorem provers as aiding tools in symbolic execution and the convenience
that declarative programming languages provide, in this thesis we attempt to introduce a strictly
declarative implementation of a theorem prover in Datalog. Our approach, namely static
declarative symbolic reasoning is implemented within the \doop{} framework for
Java Pointer Analysis, and it mainly seeks to answer \emph{"Which expressions are implied by another
expression within a program"}. The main motivation behind that decision was to leverage \doop{}'s
powerful infrastructure, and at the same time make it possible for \doop{} to utilize the reasoner
in the future for any of its analyses.
}



\abstractGr{
abstract greek
}
